---
title: "Divvy"
author: "Sungwan Kim"
date: "9/5/2017"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

This is a prediction of bicycle sharing usage in the city of Chicago for first half of 2017 using data from 2013 to 2016.


## Loading Necessary Packages

```{r}
library(tidyverse)
library(lubridate)
library(leaps)
library(tree)
library(randomForest)
library(party)
library(gbm)
```

## Importing Data
 
### Training

First we import Divvy trips for 2013 since it is formatted somewhat differently.We rename birthday variable as birthyear as it is birthyear in all other datasets. We convert starttime and stoptime into date format.
 
```{r}
trips_2013 <- read_csv("Divvy_Trips_2013.csv")
names(trips_2013)[names(trips_2013) == "birthday"] <- "birthyear"
trips_2013$starttime <- as_date(trips_2013$starttime)
trips_2013$stoptime <- as_date(trips_2013$stoptime)
```

Again, we import other datsets excluding quarter 3 and 4 in 2016 since the starttime and stoptime is formatted differently. We combine these datasets using row bind and convert dates.

```{r}
trips_2014_Q1Q2 <- read_csv("Divvy_Trips_2014_Q1Q2.csv")
trips_2014_Q3_07 <- read_csv("Divvy_Trips_2014-Q3-07.csv")
trips_2014_Q3_0809 <- read_csv("Divvy_Trips_2014-Q3-0809.csv")
trips_2014_Q4 <- read_csv("Divvy_Trips_2014-Q4.csv")
trips_2015_Q1 <- read_csv("Divvy_Trips_2015-Q1.csv")
trips_2015_Q2 <- read_csv("Divvy_Trips_2015-Q2.csv")
trips_2015_Q3_07 <- read_csv("Divvy_Trips_2015_07.csv")
trips_2015_Q3_08 <- read_csv("Divvy_Trips_2015_08.csv")
trips_2015_Q3_09 <- read_csv("Divvy_Trips_2015_09.csv")
trips_2015_Q4 <- read_csv("Divvy_Trips_2015_Q4.csv")
trips_2016_Q1 <- read_csv("Divvy_Trips_2016_Q1.csv")
trips_2016_Q2_04 <- read_csv("Divvy_Trips_2016_04.csv")
trips_2016_Q2_05 <- read_csv("Divvy_Trips_2016_05.csv")
trips_2016_Q2_06 <- read_csv("Divvy_Trips_2016_06.csv")
trips <- rbind(trips_2014_Q1Q2, trips_2014_Q3_07, trips_2014_Q3_0809, trips_2014_Q4, 
               trips_2015_Q1, trips_2015_Q2, trips_2015_Q3_07, trips_2015_Q3_08,
               trips_2015_Q3_09, trips_2015_Q4, trips_2016_Q1, trips_2016_Q2_04,
               trips_2016_Q2_05, trips_2016_Q2_06)
trips$starttime <- as_date(mdy_hm(trips$starttime))
trips$stoptime <- as_date(mdy_hm(trips$stoptime))
```

Lastly, we do appropriate date conversion for Q3 and Q4 and combined all together into one data frame.

```{r}
trips_2016_Q3 <- read_csv("Divvy_Trips_2016_Q3.csv")
trips_2016_Q4 <- read_csv("Divvy_Trips_2016_Q4.csv")
trips_2016_Q3Q4 <- rbind(trips_2016_Q3, trips_2016_Q4)
trips_2016_Q3Q4$starttime <- as_date(mdy_hms(trips_2016_Q3Q4$starttime))
trips_2016_Q3Q4$stoptime <- as_date(mdy_hms(trips_2016_Q3Q4$stoptime))
trips <- rbind(trips_2013, trips, trips_2016_Q3Q4)
```


When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
